Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures

### 2017-03-30 bn.tex
some new stuff, mainly results

\subsection{Likelihood}
\label{ssec:likelihood}

@@

\subsection{Bayesian structure learning}

@@

\subsection {Problems, solutions and problems}

@@

A natural solution to avoid parameter sensitivity of BDeu would be to
use a normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a}, i.e., to find the structure $G$ that
maximizes
\begin{equation}
P_{NML}(D;G)=\frac{P(D|\hat\theta(D;G))}{\sum_{D'}{P(D'|\hat\theta(D';G))}},
\end{equation}
where $\hat\theta$ denotes the (easy to find) maximum likelihood
parameters and the sum in the denominator goes over all the possible
$N\times n$ data matrices. While it is easy to see that this criterion
satisfies the requirement of giving equal scores to equal structures,
the normalizing constant renders the computation
infeasible. Consequently, Silander et al.~\cite{cosco.pgm08a}
suggested solving the parameter sensitivity problem by adopting the
NML-code to the column partitions, i.e., changing the Bayesian mixture
in equation~(\ref{eqn:bayesmix}) to
\begin{equation}
P^1_{NML}(D[G_i=j,i];G)=\frac{P(D|\hat\theta(D[G_i=j,i];G))}{\sum_{D'}{P(D';\hat\theta(D'|G))}},
\end{equation}
where $D'\in{\{1,\ldots,r_i\}}^{|D[G_i=j,i]|}$.  The logarithm of the
denominator is often called a regret, since it indicates the extra
code length needed compared to the code length obtained using the (a
priori unknown) maximum likelihood parameters. The regret for
$P^1_{NML}$ depends only on the length $N$ of the categorical data
vector with $r$ different categorical values.  While the naive
computation of the regret is still prohibitive, Silander et al.\
approximate it efficiently using a so-called Szpankowski
approximation~\cite{cosco.aistat03}:
\begin{eqnarray}
\label{eqn:szp1}
\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}
                               {3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
&&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}
- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
&&- \frac{r^{2} \Gamma^{2}{\left(\frac{r }{2} \right)}}
         {9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}}
+ \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
\end{eqnarray}

>>>>

\subsection{LIKELIHOOD}
\label{ssec:likelihood}

@@

\subsection{BAYESIAN STRUCTURE LEARNING}

@@

\subsection {PROBLEMS, SOLUTIONS AND PROBLEMS}

@@

A natural solution to avoid parameter sensitivity of BDeu would be to
use a normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a}, i.e., to find the structure $G$ that
maximizes
\begin{equation}
P_{NML}(D;G)=\frac{P(D|\hat\theta(D;G))}{\sum_{D'}{P(D'|\hat\theta(D';G))}},
\end{equation}
where $\hat\theta$ denotes the (easy to find) maximum likelihood
parameters and the sum in the denominator goes over all the possible
$N\times n$ data matrices. While it is easy to see that this criterion
satisfies the requirement of giving equal scores to equal structures,
the normalizing constant renders the computation
infeasible.

Consequently, Silander et al.~\cite{cosco.pgm08a}
suggested solving the BDeu parameter sensitivity problem by adopting the
NML-code to the column partitions, i.e., changing the Bayesian mixture
in equation~(\ref{eqn:bayesmix}) to
\begin{equation}
P^1_{NML}(D[G_i=j,i];G)=\frac{P(D|\hat\theta(D[G_i=j,i];G))}{\sum_{D'}{P(D';\hat\theta(D'|G))}},
\end{equation}
where $D'\in{\{1,\ldots,r_i\}}^{|D[G_i=j,i]|}$.  The logarithm of the
denominator is often called a regret, since it indicates the extra
code length needed compared to the code length obtained using the (a
priori unknown) maximum likelihood parameters. The regret for
$P^1_{NML}$ depends only on the length $N$ of the categorical data
vector with $r$ different categorical values.  While the naive
computation of the regret is still prohibitive, Silander et al.\
approximate it efficiently using a so-called Szpankowski
approximation~\cite{cosco.aistat03}:
\begin{eqnarray}
\label{eqn:szp1}
\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}
                               {3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
&&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}
- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
&&- \frac{r^{2} \Gamma^{2}{\left(\frac{r }{2} \right)}}
         {9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}}
+ \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
\end{eqnarray}

### 2017-03-30 experiments.tex
some new stuff, mainly results

To empirically compare the model selection criteria we took 20 UCI
data sets~\cite{Lichman:2013} and run 1000 train and test experiments
for 20 different sample sizes for all of them. The training was
conducted using a dynamic programming based exact structure
learning\cite{cosco.uai06} which limited the number $n$ of variables
to less than 20.

>>>>

We will next empirically compare the capacity of the qNML to that of
BIC, BDeu and fNML in identifying the data generating structures,
and producing models that are predictive and parsimonious.
It seems that none of the criteria outperform others for all
these requirements.

\subsection{FINDING GENERATING STRUCTURE}

We first compared the ability of different scoring criteria to
discover the data generating structure. For this purpose we generated
100 different 5-node Bayesian network structures with 4 edges and
another 100 structures with 7 edges.  The variables were randomly
assigned to have 2 – 4 values ($r_i \in \{2, 3, 4\}$). For each
network, we generated parameters by two different schemes. The first
scheme exactly matched the assumptions of the BDeu score with $\alpha
= 1$, i.e., the parameters were distributed by $\theta_{ij} \sim
Dir(\frac{1}{r_iq_i},\ldots,\frac{r_i}{q_i})$. The other scheme was
to generate the parameters independently from a Dirichlet distribution
$\theta_{ij} \sim Dir(\frac{1}{2},\ldots, \frac{1}{2})$ that is the
Jeffreys' prior for multinomial model. This distribution was selected
instead of the uniform distribution in order to make the generating
structure more identifiable.  For each network (structure +
parameters), we generated 100 data sets of 1000 data vectors, and
studied how different scoring criteria ranked the structure of the
generating network among all the 5-node networks as a function of
(sub)sample size.

Not surprisingly, the results indicate that when the parameter generation
mechanism matches the assumptions of the BDeu-score, the BDeu usually
also ranks the generating structure higher than the other scores
(Figure X).  On the other hand, when the paramaters are drawn from the
Jeffreys's prior, the fNML score appears to rank the generting structure
highest. This too is not surprising, since the NML distribution for the
multinomial data is known to closely match the marginal distribution of
the data when the Jefrreys' prior is used (CITE).

The underfitting tendency of BIC can also be clearly detected both for
relatively sparse networks (4 arcs) and the dense ones (7 arcs). The
ranking ability the qNML criterion appears to be between fNML and BDeu
in 3 out of for settings which is good considering that one of these
criteria is always the winner. For the dense networks with BDeu prior,
the qNML appears to perform worse than BDeu and fNML but still much
better than BIC.

\subsection{PREDICTION AND PARSIMONY}

To empirically compare the model selection criteria we took 20 UCI
data sets~\cite{Lichman:2013} and run 1000 train and test experiments
for 20 different sample sizes for all of them. The training was
conducted using a dynamic programming based exact structure
learning~\cite{cosco.uai06} which limited the number $n$ of variables
to less than 20.

### 2017-03-30 intro.tex
some new stuff, mainly results

Learning the structure of statistical dependencies can be seen as a
model selection task where each model is a different hypothesis about
the conditional dependencies between sets of variables. Traditional
model selection criteria such as the Akaike information
criterion~\cite{Akai73} and the Bayesian information
criterion~\cite{Schw78} have also been used for the task, but recent
comparisons have not been favorable for these criteria in terms of
structural stability and/or predictive
performance~\cite{cosco.pgm08a}.  The most popular criterion has been
the marginal likelihood (usually called BDeu for reasons explained
later), but recent studies~\cite{cosco.uai07,Steck08} have found this
criterion to be very sensitive to hyper parameters and to yield
undesirably complex models for small sample sizes.

The information theoretic normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a} would otherwise be an ideal candidate
for a good criterion.  but its exact calculation is likely to be
prohibitively demanding. In 2008, Silander et al. introduced a
hyper-parameter free, NML inspired criterion called a factorized NML
(fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
models without sensitivity problems.  However, from the structure
learning point of view, the fNML still sometimes appears to yield
overly complex models. In this paper we introduce another NML related
criterion, a quotient NML (qNML) that yields simpler models without
sacrificing predictive accuracy. Furthermore, unlike the fNML, the
qNML gives equal scores to the structures that encode the same
independence statements. Like other common model selection criteria,
the qNML is also consistent.

We will next briefly introduce Bayesian networks and then review the
BDeu and fNML criteria and introduce the qNML criterion.  We will also
summarize the results for 20 datasets to back up our claim of qNML
yielding parsimonious models with good predictive capabilities.

>>>>

Learning the structure of statistical dependencies can be seen as a
model selection task where each model is a different hypothesis about
the conditional dependencies between sets of variables. Traditional
model selection criteria such as the Akaike information
criterion~\cite{Akai73} and the Bayesian information
criterion~\cite{Schw78} have also been used for the task, but recent
comparisons have not been favorable for these criteria in terms of
identifying the appropriate structure and/or predictive
performance~\cite{cosco.pgm08a}. Traditionally, the most popular
criterion has been the marginal likelihood (usually called BDeu for
reasons explained later), but recent studies~\cite{cosco.uai07,Steck08}
have found this criterion to be very sensitive to hyper parameters
and to yield undesirably complex models for small sample sizes.

The information theoretic normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a} would otherwise be an ideal candidate
for a good criterion,  but its exact calculation is likely to be
prohibitively demanding. In 2008, Silander et al. introduced a
hyper-parameter free, NML inspired criterion called a factorized NML
(fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
models without sensitivity problems.  However, from the structure
learning point of view, the fNML still sometimes appears to yield
overly complex models. In this paper we introduce another NML related
criterion, a quotient NML (qNML) that yields simpler models without
sacrificing predictive accuracy. Furthermore, unlike the fNML, the
qNML gives equal scores to the structures that encode the same
independence statements. Like other common model selection criteria,
the qNML is also consistent.

We will next briefly introduce Bayesian networks and then review the
BDeu and fNML criteria and introduce the qNML criterion.  We will also
summarize the results for 20 datasets to back up our claim of qNML
yielding parsimonious models with good predictive capabilities. The
experiments with artificial data demonstrate the capability to give
high scores for the data generating model with relatively small sample
sizes.
